---
title: "Obama"
execute: 
  warning: false
  message: false
---

```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(stringr)
library(lubridate)


```

The data used in this analysis comes from the Obama Presidential Library, specifically from their Digital Research Room. It is publicly available at:

🔗 <https://www.obamalibrary.gov/digital-research-room/archived-white-house-websites-and-social-media>

The dataset is titled **`tweets.csv`**, which contains tweets posted from the official \@POTUS Twitter account during President Barack Obama’s administration. The dataset includes columns such as the text of each tweet, timestamp, and other metadata.

This analysis focuses primarily on the text column, which contains the actual content of the tweets. The goal is to examine how often President Obama mentioned key policy topics, such as healthcare, climate change, and the economy, over the course of his presidency.

```{r, echo=FALSE}


obamaa <- readr::read_csv('tweets.csv')
print(obamaa)

```

```{r, echo=FALSE}
cleanobama <- obamaa |>
  mutate(text = str_to_lower(text),
         text = str_replace_all(text, "http[[:alnum:][:punct:]]*", ""),  
         text = str_replace_all(text, "[[:punct:]]", " "),
         text = str_replace_all(text, "—@POTUS.*", ""),
         text= str_replace_all(text, "\n", "")) 


stopwords <- c("the", "and", "to", "of", "in", "a", "on", "for", "with", "is", "that","s", "at", "potus", "  ", "rt", "amp", "it","this", "are","→")


word_counts <- cleanobama |>
  mutate(words = str_split(text, "\\s+")) |>  
  unnest(words) |>  
  filter(words != "", !words %in% stopwords) |>  
  count(words, sort = TRUE)  


word_counts |>
  slice_max(n, n = 10) |> 
  ggplot(aes(x = reorder(words, n), y = n)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 10 Words in Obama's Tweets",
       x = "Word",
       y = "Frequency")

```

This graph illustrates which words Obama used most throughout his presidency in his tweets while not including "the", "and", "to", "of", "in", "a", "on", "for", "with", "is", "that","s", "at", "potus", "rt", "amp", "it","this", "are" and "→". This graph is not especially helpful as the most used words are not very descriptive of what is going on in the tweets. What we can learn is that many of his tweets are about current news as he utilizes "today" heavily and they tend to be directed to a specific audience. By utilizing "we" and "our" it illustrates that Obama is not just sharing opinions but directing messages to the general public. I do not have a clear explanation why Obama and president are utilized so heavily although I assume that Obama tended to share quotes from himself or articles written which therefore would cite him.

```{r, echo=FALSE}




policy_keywords <- c("healthcare", "education", "climate", "economy", 
                     "jobs", "tax", "immigration", "gun", "poverty")


obama <- obamaa |>
  mutate(text = str_to_lower(text),  # stringr
         text = str_replace_all(text, "[[:punct:]]", " "))  #stringr

# regular expressions
add_word_boundaries <- function(word) {
  paste0("(?<!\\w)", word, "(?!\\w)")  #look behind lookaround
}


policy_regexes <- setNames(map(policy_keywords, add_word_boundaries), policy_keywords)


policy_counts <- map_dfr(names(policy_regexes), ~{
  keyword_regex <- policy_regexes[[.x]]


  obama |>
    mutate(has_word = str_detect(text, regex(keyword_regex))) |>
    summarize(policy = .x,
              count = sum(has_word))  
}) |>
  arrange(desc(count))


print(policy_counts)


```

The number of times each of the following words is found exactly like this in unique tweets. Eliminates taxpayers and other additions to the words. I decided to create this plot because it highlighted which political themes he addressed most heavily in his tweets and is much more useful than the earlier utilized graph. The top three results are heavily tied together so I am not surprised to see them at the top as I am sure they would all appear together in each unique tweet.

```{r}
obamatime <- obamaa |>
  # Split the timestamp into date, time, timezone
  separate(timestamp, into = c("date", "time", "timezone"), sep = "\\s+") |>
  
  # Clean the time column to keep only HH:MM:SS (remove any stray characters just in case)
  mutate(time = str_replace_all(time, "[^0-9:]", "")) |>  # Regular expression cleaning
  
  # Parse time into an HMS object and create hour and time_chunk columns
  mutate(tlsime = hms(time),  
         hour_of_day = hour(tlsime),  
         time_chunk = cut(hour_of_day, breaks = seq(0, 24, by = 4), 
                          labels = c("00:00-03:59", "04:00-07:59", "08:00-11:59", 
                                     "12:00-15:59", "16:00-19:59", "20:00-23:59"), 
                          include.lowest = TRUE)) |> 
  mutate(time_chunk = factor(time_chunk, levels = c("00:00-03:59", "04:00-07:59", "08:00-11:59", 
                                                    "12:00-15:59", "16:00-19:59", "20:00-23:59"))) |>  # Fix order explicitly
  arrange(time_chunk)  # Arrange based on the time chunk order

# Now count the tweets by time of day
tweetsbytime <- obamatime |>
  count(time_chunk)  # Count tweets by time chunk

# Plot the number of tweets by time of day
tweetsbytime |>
  ggplot(aes(x = time_chunk, y = n)) +
  geom_col(fill = "steelblue") +

  labs(title = "Obama Tweets by Time of Day (4-hour Chunks)",
       x = "Time of Day (4-hour intervals)",
       y = "Number of Tweets")
```

```{r}
obamatime <- obamatime |>
  mutate(date = as.Date(date))  # Convert date to Date class if it's not already

# Extract the month and year from the date and count the tweets by month
tweetsbymonth <- obamatime |>
  mutate(month_year = format(date, "%Y-%m")) |>  # Create a column for Year-Month
  count(month_year)  # Count the tweets by month-year

# Plot the number of tweets per month
tweetsbymonth |>
  ggplot(aes(x = month_year, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Obama Tweets by Month Over His Whole Presidency",
       x = "Month-Year",
       y = "Number of Tweets") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

I was then interested if there were specific times of the year which brought upon extra tweeting from Obama but accidentally did not group by month so I had this messy graph. I was going to delete it but I found this trend quite interesting that the longer he was in office the more tweets he was sending out so I left this even though the x-axis is unclear. That jumble of numbers is the year followed by the month. For example 2012-02.

```{r}
tweetsbymonth <- obamatime |>
  mutate(month = format(date, "%m")) |>  # Extract the month (as numeric)
  count(month)  # Count the tweets by month

# Plot the number of tweets per month (grouped by month)
tweetsbymonth |>
  ggplot(aes(x = month, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Obama Tweets by Month",
       x = "Month",
       y = "Number of Tweets") +
  scale_x_discrete(labels = c("01" = "January", "02" = "February", "03" = "March", 
                              "04" = "April", "05" = "May", "06" = "June",
                              "07" = "July", "08" = "August", "09" = "September", 
                              "10" = "October", "11" = "November", "12" = "December")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

This is the graph I was actually hoping to create earlier and found that it is much less helpful in showing any real trends. I found it interesting that January had the most tweets and maybe that was due to issuing plans or goals for the new year or possibly sharing information and reports about the inauguration. But what did surprise was that it looked like a ton of tweets were being sent by him so that led me to the following calculation just out of interest.

```{r, echo=FALSE}
# Calculate the total number of tweets
total_tweets <- nrow(obamatime)

# Calculate the total number of unique days
total_days <- n_distinct(obamatime$date)

# Calculate the average number of tweets per day
average_tweets_per_day <- total_tweets / total_days

# Print the result
average_tweets_per_day



```

This function found that Obama averaged 10.68 tweets per day which seems pretty ridiculous to me although I do understand that often he was just reposting something that had been created by others. I also understand this is supposed to be an informal type of sharing information so these may be quick thoughts that he quickly types up and sends out.
